TODO Preschoolers form conventional pacts with each other to communicate about novel shapes

### 2025-01-11 write-ups/kid_tangrams_cogsci_2025/kid_tangrams_cogsci_2025.Rmd
updated text

author-information: > 
    \author{{\large \bf Veronica Boyce (vboyce@stanford.edu)} \\ Department of Psychology, \\Stanford University \And {\large \bf Bobby Sparks (TODO email) } Department of Psychology, \\Stanford University
    \AND {\large \bf Yannick Mofor (TODO email)} \\ TODO affiliation, Stanford University \And {\large \bf Michael C. Frank (mcfrank@stanford.edu)} \\ Department of Psychology, \\ Stanford University}

abstract: >
  Learning language requires not just learning the facts of a language, but also learning how to use that language to communicate. Iterated reference games require rich communicative skills as participants jointly converge on mutually understandable names for initially novel objects. Some classical experiments with young children are interpreted as showing that 4-5 year old children were incapable of succeeding at iterated references games with each other. More recent work on young children and their parents using simpler paradigms reveal greater abilities even in young children. Here, we revisit the question of young children's referential communicative abilites with iterated reference games played between pairs of 4-5 year old preschoolers. Across 51 pairs of children (TODO double check N), we find that children were Y% accurate, and that they often used descriptions similar to their partner's. Preschoolers' capacity to construct effective referring expressions in novel contexts is consistent with a larger picture of early emerging ability to use language in pragmatic and communicative ways. TODO fix wrap up sentence! TODO possibly shorten!

@@

[way to long, please chop down]
Learning language requires not just learning the facts of a language, but also learning how to use that language to communicate. One common test environment for language use is referential communication, the ability to give a description so that an interlocuter can pick out a target referent from a set of possibilities. Adults show sensitivity to both the visual context and their audience during referential communication, calibrating the type of description and the amount of information they provide to their beliefs about the interlocuters knowledge state.

One test of referential communication skill is *iterated reference games* where a describer identifies abstract shapes from an array of images repeatedly to the same partner [@clark1986; @krauss1964; @hawkins2020b; @boyce2024]. Over repetition, features of the initial verbose descriptions are conventionalized as each pair comes to agree on a shared understanding of how to label each image. Success at this task requires mastery of a number of linguistic and communicative skills including producing adequate initial descriptions, monitoring for comprehension and asking for clarification, and appropriately using the shared conversation history to inform referring expressions in later rounds.

All of these are skills that children must learn as part of their acquisition of language, but there is debate on the acquisition trajectories of different linguistic and communicative skills. Studying how children play iterated reference games can provide insight into the developmental trajectories of the ability to produce expressions to achieve joint understanding.

A study on children's referential communication declared 4-5 year old preschoolers incapable of the task of child-child referential communication [@glucksberg1966]. In their paradigm, one child was given a set of 6 blocks in a specific order. Their task was to describe the images on the block so their partner could pick out their corresponding block. As children described and selected blocks, they stacked them on pegs, so that the order of blocks on the peg was a measure of correct matching. 4-5 year old children were successful on practice trials where the blocks had drawings of farm animals and they had visual access to each others blocks; however, younger children (3-4 year olds) were not even able to understand the instructions and complete the practice trials.  On the critcal trials, the 4-5 year old children were tested with abstract shapes on the blocks and no visual access to the other child's blocks. In this condition, even after multiple rounds with the same images, children were not able to produce referential descriptions that allowed for success at the matching task.

Similar experiments with older children indicated a gradual improvement across age (up to the teen years) both at their initial accuracy and on their improvement over repetitions, although even the 9th grade sample was noticeably worse than their adult college student sample [@glucksberg1967; @krauss1969]. @glucksberg1966 attributed the young children's failures to the children’s use of idiosyncratic referring expressions unique to their own experiences, making it difficult for the pair to converge on shared descriptions. However, the complexity of the stacking paradigm and the large number of potential targets meant that this was not a clean test of children's ability to produce adequate referent expressions as the experiment had substantial task demands that could mask children's abilities.

Since then, a number of studies have targeted specific skills needed for referential communication, including expectations around consistent labeling ("conversational pacts"), awareness of other's perspectives, and sensitivity to descriptive adequacy in varying contexts. A number of these studies have found roots of understanding emerging during hte preschool years, albeit with inconsistent expression.

Young children show some sensitivity to the expectation that during cooperative communicate a person will likely re-use their own descriptions.  As listeners, preschoolers (3-5 year olds) are faster to select a target when it is referred to in a consistent way [@graham2014; @matthews2010]. However, this preference for familiar descriptions is applied regardless of whether they are said by the same person or a new person, and they sometimes verbally protesting the use of a new term ("That's a horse, not a pony!") [@matthews2010]. As the producers of descriptions, children show some partner-specificity, but are far from consistent. @koymen2014 had 4 and 6 year old children do a version of the paradigm from @brennan1996. Children used specific referring expressions ("woman's shoe") with one partner in a visual context of different types of shes. In later trials when only one shoe was present and thus a general expression like "shoe" was adequate, 6 year olds were somewhat more likely to continue using the specific term with the same partner than with a new partner, whereas 4 year olds used both specific and general expressions at the same rate regardless of partner.

Expectations of consistency are entwined with being able to take the perspective of an interlocuter. Young children are notoriously ego-centric [@epley2004; TODO there are probably better things to cite here instead], but even preschoolers are able to take other's perspectives, especially when the are not under too much cognitive load [@sanjuan2015]. Across the 2-4 year olds range, children are developing sensitivity to what is given their interlocuter based on the discourse history and visual context available to their interlocuter [@matthews2006]. Older preschoolers are more likely to use more given forms such as pronouns when their interlocuter has more context and indefinite forms when their interlocuter has less context.  4-6 year old children are more likely to use a more specific reference expression with a size modifier when their partner can see the size competitor than when their partner cannot see the competitor, but children produce the modifier inconsistently even when it is necessary to distinguish the target [@nadig2002; @nilsen2009].

The skill children seem to struggle the most with is appropriately tailoring the specificity of their utterances to the visual context. 4 year olds showing some improvement after getting specific feedback that models appropriate descriptions, but still a lot of inconsistency [@matthews2012].  Children ages 4-8 often produce the same label for a target object regardless of whether the context is close or far, leading to descriptions that are underinformative for close contexts [@leung2024]. When children must pick out one of two very similar images, 4 and 5 year olds aren't good at mentioning the relevant features, although they do better when playing in an interactive game than when not [@grigoroglou2019]. By age 5, children are sensitive to utterances that are over- or under-informative, asking for clarification on some under-informative utterances and taking longer to make selections on over-informative utterances, although children still do not ask for clarification all the time [@morisseau2013].

A common thread between many studies in referential expression production is that children perform better when the cognitive load of the task is reduced, i.e. fewer distractor referents [@abbot-smith2016]. This suggests that tailored referential expression production is possible in children, but it is likely to be masked in situations that involve cognitive load.

Since @glucksberg1966, only a few studies have revisited the question of how well children can put together different skills to communicate in an iterated reference game. @branigan2016 tested 72 8-10 year olds on iterated referential communication using a paradigm based on @wilkes-gibbs1992. In the training period, pairs of children matched tangrams in order from a pool of 8 images. Qualitatively, children on average showed the classic patterns of increasing accuracy, increasing speed, and shorter descriptions across repetitions. However, children's level of accuracy was far below what is typical for adults and varied dramatically from pair to pair.

Some of the skills required in iterated reference games are communicative rather than purely linguistic. In @bohn2019, 4 and 6 year olds communicated target images from a pool of 5 images via a video link without audio. Children's comprehension was good and increased over repetitions, and children, especially the 6 year olds,  showed signs of agreeing on conventionalized gestures with their partner.

A recent study on younger children examined children's abilities to play an iterated reference game with a parent. @leung2024 had 63 children ages 4-8 year olds play a matching game on tablets, where, on each trial, participants saw two images.  The describer's tablet had a box around one image and their task was to verbally describe the image so their partner could select the corresponding image. The target images were tangrams, with a total of 10 target images, which repeated across 4 blocks for a total of 40 trials per pair. In this task, even 4 year olds had an initial accuracy above 80% which rose to above 90% on the later repetitions.

Given the potential for task demands in @glucksberg1966 and the level of success young children had with their parents at a simplified, less demanding paradigm, we revisit the question of children's ability to converge on appropriate referential expressions with each other.
In the present study, we re-examine young children's ability to establish effective referring expressions with each other in an iterated reference game using a simplified paradigm to reduce cognitive demands.

```{r interface, fig.env = "figure*", fig.pos = "t!", out.width="\\textwidth", fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "Experimental Setup and Procedure. Panel A shows the experimental setup with the teller and guesser across the table from each other. Panel B shows the 4 possible targets; names for targets are for cross-reference with later figures only. Panel C shows the procedure for Experiment 1; within in critical block targets were ordered randomly. Panel D shows the procedure for Experiment 2. \\label{game}", cache=FALSE}

@@

## Participants
4 and 5 year old children were recruited from a university nursery school during the school day. Children played with another child from the same class. Experiment 1 was conducted between June and August 2023. Pairs of children were included in analyses if they completed at least 8 of the 12 critical trials. We had 19 complete games and 1 incomplete, but included game. Of the 40 children, 21 were girls, and the median age was 57 months, with a range of 48-70 months.

@@

[less important, could drop] As another measure of children's performance, we looked at how long each trial took to see if children were getting faster over time. We ran a Bayesian mixed effects model of how long each trial took over time: `r speed_form |> form()`. The first trial critical trial averaged `r stats_text(speed_mod_1, 1)` seconds, and children got faster over time (`r stats_text(speed_mod_1, 2)`).

@@

In the first experiment, we adapted the paradigm of @leung2024 for pairs of children, taking an already simple set up and making it shorter. Our goal was to see if young children were at all able to provide descriptions to each other adequate to identify abstract figures. so children recieved a lot of scaffolding around the experimental interaction. Sometimes, this included experimenter echoing children's descriptions which could potentially influence children's responses. In Experiment 2, we repeated the same paradigm, with tighter experimenter controls and a larger sample size.

@@

Utterances were more similar if they came from the same partnership (`r sim_across_between_mod_meta |> stats_text(2,3)`) and were slightly more similar still if they came from the same person with the partnership (`r sim_across_between_mod_meta |> stats_text(3,3)`).
Since the two expeirments had different length games, we do not want to pool them in comparing to the last block, but we can pool them comparing to the next block.

@@

Across 2 experiments and 51 pairs of 4 and 5 year old children, we tested how well children were at producing referential expressions to allow their partner to find a matching abstract shape. Children varied substantially in what sorts of descriptions they produced, but overall accuracy was high (NUMBER), indicating that children were generally able to produce adequate descriptions. Additionally, children's utterances showed signs of converging toward conceptual pacts.  While this task is substantially scaled down relative to measures used for adult competence, and children do not display all the typical trends (i.e no sign of shortening of referential expression), it does suggest that the relevant communication skills are present at least in rudimentary form by the end of the preschool years.

At least some preschoolers, in an environment with low task demands and support for understanding the structure of the task, can succeed at iterated referential communication. Our task is limited by the non-representative population of children at university nursery schools who participated, and by a limited set of target images. This set of tangram images may be easier to distinguish and refer to than some sets used with adults, leading to overall shorter utterances. We specifically targeted children's abilities to construct of referring expressions that can be jointly understood. Children were provided scaffolding around the larger coordination problems of taking turns nad talking to their partner.

In the broader picture of language acquisition, there's debate over whether children learn the facts of their language first, followed much later, but pragmatic and communicative competency (CITATIONS), or whether sensitivity to communicative intent is an early emerging skill that develops in parallel with linguistic knowledge and may bootstrap language learning (CITATIONS). The findings in the current work are most consistent with a gradual development of children's communicative and linguistic skills, where the skills emerge early and then are refined over time, as children's cogntive capacities increase. At ages 4-5 years old, children are able to use their linguistic skills communicatively, even as their utterances are not fully adult-like in form.

>>>>

author-information: > 
    \author{{\large \bf Veronica Boyce (vboyce@stanford.edu)} \\ Department of Psychology, \\Stanford University \And {\large \bf Robert Z. Sparks (bsparks@stanford.edu) } Department of Psychology, \\Stanford University
    \AND {\large \bf Yannick Mofor (TODO email)} \\ TODO affiliation, Stanford University \And {\large \bf Michael C. Frank (mcfrank@stanford.edu)} \\ Department of Psychology, \\ Stanford University}

abstract: >
   Learning language requires learning not only the content of language, but also skills that allow us to use language to communicate. Iterated reference games provide a window into such skills, requiring rich communication as participants jointly converge on mutually understandable names for initially novel objects. Some classical experiments with young children are interpreted as showing that 4-5-year old children are incapable of succeeding at iterated reference games. Here, we revisit the question of young children's referential communicative abilities using a simpler, child-friendly paradigm. Across 51 pairs of 4-5-year old children, we find that preschool-aged children were successful in an iterated reference game. Children were Y% accurate, and they often used descriptions similar to their partner's. These findings suggest that children’s capacity to construct effective referring expressions in novel contexts emerges earlier than once thought, consistent with a larger picture of the early emerging ability to pragmatically use language in communication. TODO fix wrap up sentence! TODO possibly shorten!

@@

Learning a language requires learning not only the content of that language, but also how to use the language to communicate. One test environment for language use is referential communication, the ability to give a description so that an interlocutor can pick out a target referent from a set of possibilities. Adults show sensitivity to both the visual context and their audience during referential communication, calibrating the type of description and the amount of information they provide to their beliefs about the interlocutor's knowledge state.

One test of referential communication skill is *iterated reference games* where one describes abstract shapes from an array of images repeatedly to the same partner [@clark1986; @krauss1964; @hawkins2020b; @boyce2024]. Over repetition, features of the initial verbose descriptions are conventionalized as each pair comes to agree on a shared understanding of how to label each image. Success at this task requires mastery of a number of linguistic and communicative skills, including producing adequate initial descriptions, monitoring for comprehension, asking for clarification, and appropriately using the shared conversation history to inform referring expressions in later rounds.

Indeed, even young children must make use of these skills in order to achieve success in the task. However, there is debate on the developmental trajectories of these communicative skills. Studying how children play iterated reference games can provide insight into the developmental trajectory of the ability to produce referential expressions in order to achieve joint understanding.

A study on children's referential communication suggests that 4-5-year old preschoolers struggle to succeed at child-child referential communication [@glucksberg1966]. In their paradigm, one child was given a set of 6 blocks in a specific order. Their task was to describe the images on the block so their partner could pick out the corresponding block. As children described and selected blocks, they stacked them on pegs, so that the order of blocks on the peg was a measure of correct matching. While 4-5-year old children were successful on practice trials where the blocks had drawings of farm animals and they had visual access to each other's blocks; children failed on critical trials where the blocks had drawings of abstract shapes and there was no visual access. Even after multiple rounds with the same images, children were not able to produce referential descriptions that allowed for success at the matching task. @glucksberg1966 attributed young children's failure to the children’s use of idiosyncratic referring expressions unique to their own experiences, making it difficult for the pair to converge on shared descriptions.

Similar experiments with older children indicated a gradual improvement through adolescence both at their initial accuracy and on their improvement over repetitions. Still, even the 9th grade sample was noticeably worse than the sample of adult college students [@glucksberg1967; @krauss1969]. That the task is difficult even among teenagers suggests that the complexity of the stacking paradigm and the large number of potential targets presents significant task demands that might be masking the referential communication abilities of even the youngest children.

Since then, a number of studies have revealed early emerging skills in the preschool years that support the use of referential communication. Studies show that preschool-aged children (3-5-years old) are faster to select a target when it is referred to in a consistent way [@graham2014; @matthews2010] and by 6-years more consistently produce partner-specific referential expressions [@koymen2014], suggesting that young children are sensitive to consistent use of descriptions in cooperative communication. Further, preschool-aged children adapt the informativeness of their referential expressions based on the visual content available to their interlocutor [@matthews2006;@nadig2002; @nilsen2009], showing that children at this age can reason about others’ perspectives to more efficiently communication.

While children are sensitive to others’ perspectives, they still seem to struggle with appropriately tailoring the specificity of their utterances to the visual context, resulting in utterances that are under-informative [@matthews2012; @leung2024].  When children must describe one of two very similar images, 4 and 5-year olds struggle to mention the relevant features, although they do better when playing in an interactive game than when not [@grigoroglou2019]. By 5-years old, children are sensitive to utterances that are over- or under-informative, asking for clarification on some under-informative utterances and taking longer to make selections on over-informative utterances [@morisseau2013].

A common thread between many studies in referential expression production is that children perform better when the cognitive load of the task is reduced, i.e. fewer distractor referents [@abbot-smith2016]. This paired with evidence of emerging communicative skills in young children suggests that tailored referential expression production is possible in children, but it is likely to be masked when the task demands are too high.

Since @glucksberg1966, few studies have revisited the question of whether children can successfully communicate in an iterated reference game. Recent work shows that 8-10-year olds exhibit classic patterns of increasing accuracy, increasing speed, and shorter descriptions across repetitions in an integrated reference game,  but children’s accuracy was still far below adult performance and varied dramatically between dyads [@branigan2016]. 4-6-year olds have also been shown to rely on conventionalized gestures with their partner in an integrated reference game where children could only use gestures rather than speech to communicate [@bohn2019]. A recent study with younger children from 4-8-years old examined children's abilities to play an iterated reference game with a parent using a simple, child-friendly tablet-based task. In this task, even 4-year olds had an initial accuracy above 80%, which rose to above 90% in later repetitions [@leung2024]. Together, these findings provide evidence of young children’s ability to successfully communicate in an iterated reference game under the right conditions.

Given the task demands in @glucksberg1966 and work showing that children can succeed in a simplified, less demanding paradigm, we revisit the question of preschool-aged children's ability to converge on appropriate referential expressions with each other. In the present study, we re-examine young children's ability to establish effective referring expressions with each other in an iterated reference game using a simplified tablet-based paradigm. Across 51 pairs of 4-5-year old children, we find that preschool-aged children were successful in an iterated reference game, suggesting that children’s capacity to construct effective referring expressions in novel contexts emerges earlier than once thought.

```{r interface, fig.env = "figure*", fig.pos = "t!", out.width="\\textwidth", fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "Experimental Setup and Procedure. Panel A shows the experimental setup with the teller and guesser across the table from each other. Panel B shows the 4 possible targets; names for targets are for cross-reference with later figures only. Panel C shows the procedure for Experiment 1; within critical block targets were ordered randomly. Panel D shows the procedure for Experiment 2. \\label{game}", cache=FALSE}

@@

## Participants
4 and 5-year old children were recruited from a university nursery school during the school day. Children played with another child from the same class. Experiment 1 was conducted between June and August 2023. Pairs of children were included in analyses if they completed at least 8 of the 12 critical trials. We had 19 complete games and 1 incomplete, but included the game. Of the 40 children, 21 were girls, and the median age was 57 months, with a range of 48-70 months.

@@

[less important, could drop] As another measure of children's performance, we looked at how long each trial took to see if children were getting faster over time. We ran a Bayesian mixed effects model of how long each trial took over time: `r speed_form |> form()`. The first critical trial averaged `r stats_text(speed_mod_1, 1)` seconds, and children got faster over time (`r stats_text(speed_mod_1, 2)`).

@@

In the first experiment, we adapted the paradigm of @leung2024 for pairs of children, taking an already simple set up and making it shorter. Our goal was to see if young children were at all able to provide descriptions to each other adequate to identify abstract figures. so children received a lot of scaffolding around the experimental interaction. Sometimes, this included experimenters echoing children's descriptions which could potentially influence children's responses. In Experiment 2, we repeated the same paradigm, with tighter experimenter controls and a larger sample size.

@@

Utterances were more similar if they came from the same partnership (`r sim_across_between_mod_meta |> stats_text(2,3)`) and were slightly more similar still if they came from the same person with the partnership (`r sim_across_between_mod_meta |> stats_text(3,3)`).
Since the two experiments had different length games, we do not want to pool them in comparing to the last block, but we can pool them comparing to the next block.

@@

Across 2 experiments and 51 pairs of 4 and 5-year old children, we tested how well children were at producing referential expressions to allow their partner to find a matching abstract shape. Children varied substantially in what sorts of descriptions they produced, but overall accuracy was high (NUMBER), indicating that children were generally able to produce adequate descriptions. Additionally, children's utterances showed signs of converging toward conceptual pacts.  While this task is substantially scaled down relative to measures used for adult competence, and children do not display all the typical trends (i.e no sign of shortening of referential expression), it does suggest that the relevant communication skills are present at least in rudimentary form by the end of the preschool years.

At least some preschoolers, in an environment with low task demands and support for understanding the structure of the task, can succeed at iterated referential communication. Our task is limited by the non-representative population of children at university nursery schools who participated, and by a limited set of target images. This set of tangram images may be easier to distinguish and refer to than some sets used with adults, leading to overall shorter utterances. We specifically targeted children's abilities to construct referring expressions that can be jointly understood. Children were provided scaffolding around the larger coordination problems of taking turns and talking to their partner.

In the broader picture of language acquisition, there's debate over whether children learn the facts of their language first, followed much later, by pragmatic and communicative competency (CITATIONS), or whether sensitivity to communicative intent is an early emerging skill that develops in parallel with linguistic knowledge and may bootstrap language learning (CITATIONS). The findings in the current work are most consistent with a gradual development of children's communicative and linguistic skills, where the skills emerge early and then are refined over time, as children's cognitive capacities increase. At 4-5-years old, children are able to use their linguistic skills communicatively, even as their utterances are not fully adult-like in form.
